\LARGE{\bf{Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})}}\footnotemark

### 2014-09-04 ResponseToReview.tex

Nevertheless, the manuscript is worth publishing as it does present data that may help rule out some possible explanations as to why tb1 is so common in parviglumus.\\

\res{Thank you; we hope that our paper in conjunction with future research and experiments can help shed light on the ecological significance of both \emph{tb1} and the \emph{Hopscotch} insertion in teosinte.}

6) This paper is fairly maize-centric. A more general message is included in the discussion, not as much in the introduction, and not at all in the abstract. This could be easily remedied.\\

\lev{Do you guys think what we tweaked in the abstract and the discussion good enough to have addressed this point?}

@@

Methods and statistics were appropriately used as far as my expertise allows me to judge.
Specific suggestions follow:

7) Introduction:
Page 4, Lines 7 - 9: You say ``Additionally, many domesticates show reduced genetic diversity when compared to their wild progenitors, and an understanding of the distribution of diversity in the wild and its phenotypic effects has become increasingly useful to crop improvement \citep{KovachMcCouch2008}.".  Can you tell us why an understanding is useful? Can you give some examples?\\

\res{We have expanded on this concept in the introduction on page 4, lines 24-45: ``Domesticated crops and their wild progenitors provide an excellent system in which to study adaptation and genomic changes associated with human-mediated selection \citep{Ross-Ibarra2007}. 
Plant domestication usually involves a suite of phenotypic changes such as loss of seed shattering and increased fruit or grain size, which are commonly referred to as `domestication syndrome' \citep{OlsenWendel2013}, and much of the study of domestication has focused on understanding the genetic variation underlying these traits \citep{OlsenGross2010}. 
Because most domesticates show reduced genetic diversity relative to their wild counterparts, effort has been made to identify agronomically useful variation in crop wild relatives \citep{flint2009wide}. Often, after identification, the alleles conferring these beneficial traits are bred into domesticates for crop improvement. For example, \emph{Oryza rufipogon}, the wild progenitor of domesticated rice, has proven useful for the integration of a number of beneficial QTL controlling traits such as grain size and yield into domesticated rice \citep{KovackMcCouch2008}. In addition to researching the role of wild alleles in domesticates, researchers have also investigated the role of variation in domesticated taxa in the evolution of feral and weedy populations \citep{ellstrand2010crops}. A recent paper examining the genomic changes associated with rabbit domestication found that many domestication phenotypes are due to mutations in many small effect loci, and that many of these mutations are still found segregating in natural populations of wild rabbit, suggesting that selection acting on standing variation \citep{Rabbit}. There are numerous examples of weedy plants evolving to mimic their domestic counterpart, and even some instances where weedy plants have evolved domesticates (reviewed in \citealt{ellstrand2010crops}). But even though domesticated alleles are often found segregating in wild relatives \citep{Gallavotti2004, SigmonVollbrecht2010}, we know almost nothing about the ecological role of this variation in natural populations. In this paper we present an ecological genetic analysis of the domestication locus \emph{tb1} in natural populations of the wild ancestor of domesticated maize."}

### 2014-09-07 manuscript.tex
edits
to response, text, line/page #s

\emph{Methods:} Prevalence of the \emph{Hopscotch} element was assessed across an Americas-wide sample of 1110 maize and teosinte individuals using a co-dominant PCR assay.  Population genetic summaries were calculated for a subset of individuals from four teosinte populations in central Mexico. Phenotypic data were also collected using seed from a single teosinte population where \emph{Hopscotch} was found segregating at high frequency.

@@

Domesticated crops and their wild progenitors provide an excellent system in which to study adaptation and genomic changes associated with human-mediated selection \citep{Ross-Ibarra2007}. 
Plant domestication usually involves a suite of phenotypic changes such as loss of seed shattering and increased fruit or grain size, which are commonly referred to as `domestication syndrome' \citep{OlsenWendel2013}, and much of the study of domestication has focused on understanding the genetic variation underlying these traits \citep{OlsenGross2010}. 
Because most domesticates show reduced genetic diversity relative to their wild counterparts, effort has been made to identify agronomically useful variation in crop wild relatives \citep{flint2009wide}. Often, after identification, the alleles conferring these beneficial traits are bred into domesticates for crop improvement. For example, \emph{Oryza rufipogon}, the wild progenitor of domesticated rice, has proven useful for the integration of a number of beneficial QTL controlling traits such as grain size and yield into domesticated rice \citep{KovackMcCouch2008}. In addition to researching the role of wild alleles in domesticates, researchers have also investigated the role of variation in domesticated taxa in the evolution of feral and weedy populations \citep{ellstrand2010crops}. A recent paper examining the genomic changes associated with rabbit domestication found that many domestication phenotypes are due to mutations in many small effect loci, and that many of these mutations are still found segregating in natural populations of wild rabbit, suggesting that selection acting on standing variation \citep{Rabbit}. There are numerous examples of weedy plants evolving to mimic their domestic counterpart, and even some instances where weedy plants have evolved domesticates (reviewed in \citealt{ellstrand2010crops}). But even though domesticated alleles are often found segregating in wild relatives \citep{Gallavotti2004, SigmonVollbrecht2010}, we know almost nothing about the ecological role of this variation in natural populations. In this paper we present an ecological genetic analysis of the domestication locus \emph{tb1} in natural populations of the wild ancestor of domesticated maize. 
%some effort has been invested into identifying agronomically useful loci in wild relatives \citep{}understanding how alleles from wild relatives behave in domesticated crops \citep[e.g.][]{Bai2007}, or identifying agronomically usefulvery little is known about the role that alleles found most commonly in domesticates play in natural populations of their wild progenitors \citep{Whitton1997}.
%In addition to the loss of diversity due to selection on loci important for domestication, demography and the effects of linked selection have acted to decrease diversity genome-wide, leading to lower genetic diversity in many domesticates relative to their wild progenitors \citep{KovachMcCouch2008}.

%Many of these wild alleles no longer found in their domestic counterparts often harbor beneficial traits such as disease resistance or drought tolerance, and an understanding of the distribution of this diversity in the wild and its phenotypic effects has become increasingly useful to crop improvement \citep{KovachMcCouch2008}. 
%For example, \emph{Oryza rufipogon}, the wild progenitor of domesticated rice, has proven useful for the integration of a number of beneficial QTL controlling traits such as grain size and yield into domesticated rice \citep{KovackMcCouch2008}. 
%\jri{why not include some specific teosinte examples like those cited in Hufford 2012 TIG?} Furthermore, in teosinte, many of the alleles underlying domestication phenotypes have been shown to be segregating in populations of teosinte, suggesting that natural variation in teosinte has played an important role in the domestication of maize \citep{Gallavotti2004, SigmonVollbrect2010}.

@@

\textbf{Genotyping analysis---}To calculate differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) we used HierFstat \citep{Goudet2005}. These analyses only included populations ($n = 32$) in which eight or more chromosomes were sampled. To test the hypothesis that the \emph{Hopscotch} insertion may be adaptive under certain environmental conditions, we looked for significant associations between \emph{Hopscotch} frequency and environmental variables using the software BayEnv \citep{Coop2010}. BayEnv creates a covariance matrix of relatedness between populations and then tests a null model that allele frequencies in populations are determined by the covariance matrix of relatedness alone against the alternative model that allele frequencies are determined by a combination of the covariance matrix and an environmental variable, producing a posterior probability (\emph{i.e.,} Bayes Factor; \citealt{Coop2010}). We used teosinte (ssp. \emph{parviglumis} and ssp. \emph{mexicana} genotyping and covariance data from \citet{Pyhajarvi2013} for BayEnv, with the \emph{Hopscotch} insertion coded as an additional SNP. SNP data from \citep{Pyhajarvi2013} were obtained using the MaizeSNP50 BeadChip and Infinium HD Assay (Illumina, San Diego, CA, USA). Environmental data were obtained from \url{www.worldclim.org} and soil data were downloaded from the Harmonized World Soil Database \citep{FAOHWSD} at \url{www.harvestchoice.org}.  These data represent average values for the last several decades (climatic data) or are likely stable over time (soil data) and therefore represent conditions important for local adaptation of our samples.  Information from these data sets was summarized by principle component analysis following \citet{Pyhajarvi2013}.

@@

Tillering index data for each genotypic class did not meet the criteria for a repeated measures ANOVA, so we transformed the data with a Box-Cox transformation ($\lambda=0$) in the Car Package for R \citep{FoxWeisberg2011} to improve the normality and homogeneity of variance among genotype classes.  We analyzed relationships between genotype and tillering index and tiller number using a repeated measures ANOVA through a general linear model function implemented in SAS v.9.3 (SAS Institute Inc., Cary, NC, USA). Additionally, in order to compare any association between \emph{Hopscotch} genotype and tillering and associations at other presumably unrelated traits, we performed an ANOVA between culm diameter and genotype using the same general linear model in SAS. Culm diameter is not believed to be correlated with tillering index or variation at \emph{tb1} and is used as our independent trait for phenotyping analyses. SAS code used for analysis is available at \url{XXXX}.

@@

\textbf{Genotyping---}The genotype at the \emph{Hopscotch} insertion was confirmed with two PCRs for 837 individuals. \jri{does this conflict with the 1100 mentioned in the summary?} \mbh{So for the rest of the samples, did we have only one PCR that worked or no amplification?  We should clarify here the number for which we obtained \emph{Hopscotch} data and also use this number in the abstract.  In the methods we can keep the total number we screened (1100).}
Among the 247 maize landrace accessions genotyped, all but eight were homozygous for the presence of the insertion (Appendix 1 and Appendix 2, See Supplemental Materials with the online version of this article). Within our \emph{parviglumis} and \emph{mexicana} samples we found the \emph{Hopscotch} insertion segregating in in 37 ($n=86$) and four ($n=17$) populations, respectively, and at highest frequency in the states of Jalisco, Colima, and Michoac\'{a}n in central-western Mexico (Fig. \ref{Fig1Map}). Using our \emph{Hopscotch} genotyping, we calculated differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) for populations in which we sampled eight or more individuals. We found that F\textsubscript{CT} = 0, and levels of F\textsubscript{ST} among populations within each subspecies (0.22) and among all populations (0.23) are similar to those reported genome-wide in previous studies (\citealt{Pyhajarvi2013}; Table \ref{Table1Fst}). Although we found large variation in \emph{Hopscotch} allele frequency among our populations, BayEnv analysis did not indicate a correlation between the \emph{Hopscotch} insertion and environmental variables (all Bayes Factors $<$ 1).

\textbf{Sequencing---}To investigate patterns of sequence diversity and linkage disequilibrium (LD) in the \emph{tb1} region, we sequenced two small ($<$1kb) regions upstream of the \emph{tb1} ORF in four populations. After alignment and singleton checking we recovered 48 and 40 segregating sites for the 5' UTR region (Region 1) and the 66kb upstream region (Region 2), respectively. For Region 1, Ejutla A has the highest values of haplotype diversity and $\theta_\pi$, while Ejutla B and La Mesa have comparable values of these summary statistics, and San Lorenzo has much lower values. Additionally, Tajima's D is strongly negative in the two Ejutla populations and La Mesa, but is less negative in San Lorenzo (Table \ref{Table2Diversity}, Appendix 2, See Supplemental Materials with the online version of this article). For Region 2, haplotype diversity and $\theta_\pi$, are similar for Ejutla A and Ejutla B, while La Mesa and San Lorenzo have slightly lower values for these statistics (Table \ref{Table2Diversity}). Tajima's D is positive in all populations except La Mesa, \jri{positive D indicates too few rare variants. the discussion also says D is negative everywhere.  is D positive or is the text wrong? please check that discussion and results are consistent and both match what is in the table.} \mbh{based on my read of the manuscript, the text is correct here (and it agrees with Table 2 and the Discussion)...if D is negative in La Mesa, that indicates an excess of rare variants in this population; also, the consistently negative D comment in the Discussion is referring to the 5' region of tb1 which shows negative D across populations so this is correct; check me on this, though, LV} indicating an excess of low frequency variants in this population (Table \ref{Table2Diversity}). Pairwise values of F\textsubscript{ST} within population pairs Ejutla A/Ejutla B and San Lorenzo/La Mesa are close to zero for both sequenced regions as well as for the \emph{Hopscotch}, while they are high for other population pairs (Table \ref{Table1Fst}). Neighbor joining trees of our sequence data and data from the teosinte inbred lines (TILs; data from Maize HapMapV2, \citealt{Chia2012}) do not reveal any clear clustering pattern with respect to population or \emph{Hopscotch} genotype (Appendix 5, See Supplemental Materials with the online version of this article); individuals within our sample that have the \emph{Hopscotch} insertion do not group with the teosinte inbred lines or domesticated maize that have the \emph{Hopscotch} insertion.

@@

Remaining explanations for differential frequencies of the \emph{Hopscotch} among teosinte populations include both genetic drift and natural selection.  Previous studies using both SSRs and genome-wide SNP data have found evidence for a population bottleneck in the San Lorenzo population \citep{Hufford2010, Pyhajarvi2013}, and the lower levels of sequence diversity in this population in the 5' UTR (Region 1) coupled with more positive values of Tajima's D are consistent with these earlier findings. Such population bottlenecks can exaggerate the effects of genetic drift through which the \emph{Hopscotch} allele may have risen to high frequency entirely by chance. A bottleneck in San Lorenzo, however, does not explain the high frequency of the \emph{Hopscotch} in multiple populations in the Jalisco cluster. Moreover, available information on diversity and population structure among Jaliscan populations \citep{Hufford2010, Pyhajarvi2013} is not suggestive of recent colonization or other demographic events that would predict a high frequency of the allele across populations. 
Finally, diversity values in the 5' UTR of \emph{tb1} are suggestive of natural selection acting upon the gene in natural populations of \emph{parviglumis}. Overall nucleotide diversity is 76\% less than seen in the sequences from the 66kb upstream region, and Tajima's D is considerably lower and consistently negative (Table XXX) \jri{please add table}. In fact, values of Tajima's D in the 5' UTR are toward the extreme negative end of the distribution of this statistic previously calculated across loci sequenced in \emph{parviglumis} \citep{Wright27052005, Moeller01072007}. Though not definitive, these results are consistent with the action of selection on the upstream region of \emph{tb1}, perhaps suggesting an ecological role for the gene in \emph{parviglumis}.

\jri{matt, do we inclue mention of weber 2007 in the introduction as part of the motivation? might be good?} Significant effects of the \emph{Hopscotch} insertion on lateral branch length, number of cupules, and tillering index in domesticated maize have been well documented \citep{Studer2011}. \citet{Weber2007} described significant phenotypic associations between markers in and around \emph{tb1} and lateral branch length and female ear length in a sample from 74 natural populations of \emph{parviglumis} \citep{Weber2007}; however, these data did not include markers from the \emph{Hopscotch} region 66kb upstream of \emph{tb1}. Our study is the first to explicitly examine the phenotypic effects of the \emph{Hopscotch} insertion across a wide collection of individuals sampled from natural populations of teosinte. We have found no significant effect of the \emph{Hopscotch} insertion on tillering index or tiller number, a result that is discordant with its clear phenotypic effects in maize. One interpretation of this result would be that the \emph{Hopscotch} controls tillering in maize \citep{Studer2011}, but tillering in teosinte is affected by variation at other loci. Consistent with this interpretation, \emph{tb1} is thought to be part of a complex pathway controlling branching, tillering and other phenotypic traits \citep{KebromBrutnell2007, Clark2006}.

A recent study by \citet{StuderDoebley2012} examined variation across traits in a three-taxa allelic series at the \emph{tb1} locus. \citet{StuderDoebley2012} introgressed nine unique teosinte \emph{tb1} segments (one from \emph{Zea diploperennis}, and four each from \emph{mexicana} and \emph{parviglumis}) into an inbred maize background and investigated their phenotypic effects. Phenotypes were shown to cluster by taxon, indicating \emph{tb1} may underlie morphological diversification of \emph{Zea}. Additional analysis in \citet{StuderDoebley2012} suggested tillering index was controlled both by \emph{tb1} and loci elsewhere in the genome. Clues to the identity of these loci may be found in QTL studies that have identified loci controlling branching architecture (\emph{e.g.,} \citealt{DoebleyStec1991, DoebleyStec1993}). Many of these loci (\emph{grassy tillers}, \emph{gt1}; \emph{tassel-replaces-upper-ears1}, \emph{tru1}; \emph{terminal ear1}, \emph{te1}) have been shown to interact with \emph{tb1} \citep{Whipple2011, Li2012},  and both \emph{tru1} and \emph{te1} affect the same phenotypic traits as \emph{tb1} \citep{DoebleyStecGustus1995}. \emph{tru1}, for example, has been shown to act either epistatically or downstream of \emph{tb1}, affecting both branching architecture (decreased apical dominance) and tassel phenotypes (shortened tassel and shank length and reduced tassel number; \citealt{Li2012}). Variation in these additional loci may have affected tillering in our collections and contributed to the lack of correlation we see between \emph{Hopscotch} genotype and tillering. 
\jri{do we mention photoperiod as a possible explanation for phenotyping 1 results?}
%Finally, although photoperiod for Phenotyping 2 reasonably approximated that of the normal \emph{parviglumis} growing season, greenhouse-specific environmental conditions (plant density, light regime, etc...) may have contributed to tillering responses different from those found in nature, obscuring the effect of the \emph{Hopscotch} insertion on variation.

In conclusion, our findings demonstrate that the \emph{Hopscotch} allele is widespread in populations of \emph{parviglumis} and \emph{mexicana} and occasionally at high allele frequencies. 
Analysis of linkage using SNPs from across chromosome 1 does not suggest that the \emph{Hopscotch} allele is present in these populations due to recent introgression, and it seems unlikely that the insertion would have drifted to high frequency in multiple populations. 
We do, however, find preliminary evidence of selection on the \emph{tb1} locus in \emph{parviglumis}.  Coupled with our observation of high frequency of the \emph{Hopscotch} insertion in a number of populations, this suggests that the locus may play an ecological role in teosinte. 
In contrast to domesticated maize, the \emph{Hopscotch} insertion does not appear to have a large effect on tillering in a diverse sample of \emph{parviglumis} from a natural population and the phenotypic consequences of variation at \emph{tb1} thus remain unclear. Future studies should examine expression levels of \emph{tb1} in teosinte with and without the \emph{Hopscotch} insertion and further characterize the effects of additional loci involved in branching architecture (e.g. \emph{gt1}, \emph{tru1}, and \emph{te1}). These data, in conjunction with more exhaustive phenotyping, should help to further clarify the ecological significance of the domesticated \emph{tb1} allele in natural populations of teosinte.

>>>>

\emph{Methods:} Prevalence of the \emph{Hopscotch} element was assessed across an Americas-wide sample of 837 maize and teosinte individuals using a co-dominant PCR assay.  Population genetic summaries were calculated for a subset of individuals from four teosinte populations in central Mexico. Phenotypic data were also collected using seed from a single teosinte population where \emph{Hopscotch} was found segregating at high frequency.

@@

Domesticated crops and their wild progenitors provide an excellent system in which to study adaptation and genomic changes associated with human-mediated selection \citep{Ross-Ibarra2007}. 
Plant domestication usually involves a suite of phenotypic changes such as loss of seed shattering and increased fruit or grain size, which are commonly referred to as `domestication syndrome' \citep{OlsenWendel2013}, and much of the study of domestication has focused on understanding the genetic variation underlying these traits \citep{OlsenGross2010}. 
Because most domesticates show reduced genetic diversity relative to their wild counterparts, effort has been made to identify agronomically useful variation in crop wild relatives \citep{flint2009wide}. Often, after identification, the alleles conferring these beneficial traits are bred into domesticates for crop improvement. For example, \emph{Oryza rufipogon}, the wild progenitor of domesticated rice, has proven useful for the integration of a number of beneficial QTL controlling traits such as grain size and yield into domesticated rice \citep{KovackMcCouch2008}. In addition to researching the role of wild alleles in domesticates, researchers have also investigated the role of variation in domesticated taxa in the evolution of feral and weedy populations \citep{ellstrand2010crops}. A recent paper examining the genomic changes associated with rabbit domestication found that many domestication phenotypes are due to mutations in many small effect loci, and that many of these mutations are still found segregating in natural populations of wild rabbit, suggesting that selection acting on standing variation \citep{Rabbit}. There are numerous examples of weedy plants evolving to mimic their domestic counterpart, and even some instances where weedy plants have evolved domesticates (reviewed in \citealt{ellstrand2010crops}). But even though domesticated alleles are often found segregating in wild relatives \citep{Gallavotti2004, SigmonVollbrecht2010}, we know almost nothing about the ecological role of this variation in natural populations. In this paper we present an ecological genetic analysis of the domestication locus \emph{tb1} in natural populations of the wild ancestor of domesticated maize.

@@

\textbf{Genotyping analysis---}To calculate differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) we used HierFstat \citep{Goudet2005}. These analyses only included populations ($n = 32$) in which eight or more chromosomes were sampled. To test the hypothesis that the \emph{Hopscotch} insertion may be adaptive under certain environmental conditions, we looked for significant associations between \emph{Hopscotch} frequency and environmental variables using the software BayEnv \citep{Coop2010}. BayEnv creates a covariance matrix of relatedness between populations and then tests a null model that allele frequencies in populations are determined by the covariance matrix of relatedness alone against the alternative model that allele frequencies are determined by a combination of the covariance matrix and an environmental variable, producing a posterior probability (\emph{i.e.,} Bayes Factor; \citealt{Coop2010}). We used teosinte (ssp. \emph{parviglumis} and ssp. \emph{mexicana}) genotyping and covariance data from \citet{Pyhajarvi2013} for BayEnv, with the \emph{Hopscotch} insertion coded as an additional SNP. SNP data from \citep{Pyhajarvi2013} were obtained using the MaizeSNP50 BeadChip and Infinium HD Assay (Illumina, San Diego, CA, USA). Environmental data were obtained from \url{www.worldclim.org} and soil data were downloaded from the Harmonized World Soil Database \citep{FAOHWSD} at \url{www.harvestchoice.org}.  These data represent average values for the last several decades (climatic data) or are likely stable over time (soil data) and therefore represent conditions important for local adaptation of our samples.  Information from these data sets was summarized by principle component analysis following \citet{Pyhajarvi2013}.

@@

Tillering index data for each genotypic class did not meet the criteria for a repeated measures ANOVA, so we transformed the data with a Box-Cox transformation ($\lambda=0$) in the Car Package for R \citep{FoxWeisberg2011} to improve the normality and homogeneity of variance among genotype classes.  We analyzed relationships between genotype and tillering index and tiller number using a repeated measures ANOVA through a general linear model function implemented in SAS v.9.3 (SAS Institute Inc., Cary, NC, USA). Additionally, in order to compare any association between \emph{Hopscotch} genotype and tillering and associations at other presumably unrelated traits, we performed an ANOVA between culm diameter and genotype using the same general linear model in SAS. Culm diameter is not believed to be correlated with tillering index or variation at \emph{tb1} and is used as our independent trait for phenotyping analyses. SAS code used for analysis is available at \lev{\url{XXXX}}.

@@

\textbf{Genotyping---}The genotype at the \emph{Hopscotch} insertion was confirmed with two PCRs for 837 individuals of the 1100 screened. Among the 247 maize landrace accessions genotyped, all but eight were homozygous for the presence of the insertion (Appendix 1 and Appendix 2, See Supplemental Materials with the online version of this article). Within our \emph{parviglumis} and \emph{mexicana} samples we found the \emph{Hopscotch} insertion segregating in in 37 ($n=86$) and four ($n=17$) populations, respectively, and at highest frequency in the states of Jalisco, Colima, and Michoac\'{a}n in central-western Mexico (Fig. \ref{Fig1Map}). Using our \emph{Hopscotch} genotyping, we calculated differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) for populations in which we sampled eight or more individuals. We found that F\textsubscript{CT} = 0, and levels of F\textsubscript{ST} among populations within each subspecies (0.22) and among all populations (0.23) are similar to those reported genome-wide in previous studies (\citealt{Pyhajarvi2013}; Table \ref{Table1Fst}). Although we found large variation in \emph{Hopscotch} allele frequency among our populations, BayEnv analysis did not indicate a correlation between the \emph{Hopscotch} insertion and environmental variables (all Bayes Factors $<$ 1).

\textbf{Sequencing---}To investigate patterns of sequence diversity and linkage disequilibrium (LD) in the \emph{tb1} region, we sequenced two small ($<$1kb) regions upstream of the \emph{tb1} ORF in four populations. After alignment and singleton checking we recovered 48 and 40 segregating sites for the 5' UTR region (Region 1) and the 66kb upstream region (Region 2), respectively. For Region 1, Ejutla A has the highest values of haplotype diversity and $\theta_\pi$, while Ejutla B and La Mesa have comparable values of these summary statistics, and San Lorenzo has much lower values. Additionally, Tajima's D is strongly negative in the two Ejutla populations and La Mesa, but is less negative in San Lorenzo (Table \ref{Table2Diversity}, Appendix 2, See Supplemental Materials with the online version of this article). For Region 2, haplotype diversity and $\theta_\pi$, are similar for Ejutla A and Ejutla B, while La Mesa and San Lorenzo have slightly lower values for these statistics (Table \ref{Table2Diversity}). Tajima's D is positive in all populations except La Mesa, indicating an excess of low frequency variants in this population (Table \ref{Table2Diversity}). Pairwise values of F\textsubscript{ST} within population pairs Ejutla A/Ejutla B and San Lorenzo/La Mesa are close to zero for both sequenced regions as well as for the \emph{Hopscotch}, while they are high for other population pairs (Table \ref{Table1Fst}). Neighbor joining trees of our sequence data and data from the teosinte inbred lines (TILs; data from Maize HapMapV2, \citealt{Chia2012}) do not reveal any clear clustering pattern with respect to population or \emph{Hopscotch} genotype (Appendix 5, See Supplemental Materials with the online version of this article); individuals within our sample that have the \emph{Hopscotch} insertion do not group with the teosinte inbred lines or domesticated maize that have the \emph{Hopscotch} insertion.

@@

Remaining explanations for differential frequencies of the \emph{Hopscotch} among teosinte populations include both genetic drift and natural selection.  Previous studies using both SSRs and genome-wide SNP data have found evidence for a population bottleneck in the San Lorenzo population \citep{Hufford2010, Pyhajarvi2013}, and the lower levels of sequence diversity in this population in the 5' UTR (Region 1) coupled with more positive values of Tajima's D are consistent with these earlier findings. Such population bottlenecks can exaggerate the effects of genetic drift through which the \emph{Hopscotch} allele may have risen to high frequency entirely by chance. A bottleneck in San Lorenzo, however, does not explain the high frequency of the \emph{Hopscotch} in multiple populations in the Jalisco cluster. Moreover, available information on diversity and population structure among Jaliscan populations \citep{Hufford2010, Pyhajarvi2013} is not suggestive of recent colonization or other demographic events that would predict a high frequency of the allele across populations. 
Finally, diversity values in the 5' UTR of \emph{tb1} are suggestive of natural selection acting upon the gene in populations of \emph{parviglumis}. Overall nucleotide diversity is 76\% less than seen in the sequences from the 66kb upstream region, and Tajima's D is considerably lower and consistently negative across populations (Table \ref{Table2Diversity}). In fact, values of Tajima's D in the 5' UTR are toward the extreme negative end of the distribution of this statistic previously calculated across loci sequenced in \emph{parviglumis} \citep{Wright27052005, Moeller01072007}. Though not definitive, these results are consistent with the action of selection on the upstream region of \emph{tb1}, perhaps suggesting an ecological role for the gene in \emph{parviglumis}.

Significant effects of the \emph{Hopscotch} insertion on lateral branch length, number of cupules, and tillering index in domesticated maize have been well documented \citep{Studer2011}. \citet{Weber2007} described significant phenotypic associations between markers in and around \emph{tb1} and lateral branch length and female ear length in a sample from 74 natural populations of \emph{parviglumis} \citep{Weber2007}; however, these data did not include markers from the \emph{Hopscotch} region 66kb upstream of \emph{tb1}. Our study is the first to explicitly examine the phenotypic effects of the \emph{Hopscotch} insertion across a wide collection of individuals sampled from natural populations of teosinte. We have found no significant effect of the \emph{Hopscotch} insertion on tillering index or tiller number, a result that is discordant with its clear phenotypic effects in maize. One interpretation of this result would be that the \emph{Hopscotch} controls tillering in maize \citep{Studer2011}, but tillering in teosinte is affected by variation at other loci. Consistent with this interpretation, \emph{tb1} is thought to be part of a complex pathway controlling branching, tillering and other phenotypic traits \citep{KebromBrutnell2007, Clark2006}.

A recent study by \citet{StuderDoebley2012} examined variation across traits in a three-taxa allelic series at the \emph{tb1} locus. \citet{StuderDoebley2012} introgressed nine unique teosinte \emph{tb1} segments (one from \emph{Zea diploperennis}, and four each from \emph{mexicana} and \emph{parviglumis}) into an inbred maize background and investigated their phenotypic effects. Phenotypes were shown to cluster by taxon, indicating \emph{tb1} may underlie morphological diversification of \emph{Zea}. Additional analysis in \citet{StuderDoebley2012} suggested tillering index was controlled both by \emph{tb1} and loci elsewhere in the genome. Clues to the identity of these loci may be found in QTL studies that have identified loci controlling branching architecture (\emph{e.g.,} \citealt{DoebleyStec1991, DoebleyStec1993}). Many of these loci (\emph{grassy tillers}, \emph{gt1}; \emph{tassel-replaces-upper-ears1}, \emph{tru1}; \emph{terminal ear1}, \emph{te1}) have been shown to interact with \emph{tb1} \citep{Whipple2011, Li2012},  and both \emph{tru1} and \emph{te1} affect the same phenotypic traits as \emph{tb1} \citep{DoebleyStecGustus1995}. \emph{tru1}, for example, has been shown to act either epistatically or downstream of \emph{tb1}, affecting both branching architecture (decreased apical dominance) and tassel phenotypes (shortened tassel and shank length and reduced tassel number; \citealt{Li2012}). Variation in these additional loci may have affected tillering in our collections and contributed to the lack of correlation we see between \emph{Hopscotch} genotype and tillering.

In conclusion, our findings demonstrate that the \emph{Hopscotch} allele is widespread in populations of \emph{parviglumis} and \emph{mexicana} and occasionally at high allele frequencies. 
Analysis of linkage using SNPs from across chromosome 1 does not suggest that the \emph{Hopscotch} allele is present in these populations due to recent introgression, and it seems unlikely that the insertion would have drifted to high frequency in multiple populations. 
We do, however, find preliminary evidence of selection on the \emph{tb1} locus in \emph{parviglumis}.  Coupled with our observation of high frequency of the \emph{Hopscotch} insertion in a number of populations, this suggests that the locus may play an ecological role in teosinte.

In contrast to domesticated maize, the \emph{Hopscotch} insertion does not appear to have a large effect on tillering in a diverse sample of \emph{parviglumis} from a natural population and the phenotypic consequences of variation at \emph{tb1} thus remain unclear. Future studies should examine expression levels of \emph{tb1} in teosinte with and without the \emph{Hopscotch} insertion and further characterize the effects of additional loci involved in branching architecture (e.g. \emph{gt1}, \emph{tru1}, and \emph{te1}). These data, in conjunction with more exhaustive phenotyping, should help to further clarify the ecological significance of the domesticated \emph{tb1} allele in natural populations of teosinte.
